Reformulating phylogenetic mixed models to improve flexibility and speed

### 2020-02-07 phyloglmm_ms.tex
pushing edits, I think we are done

The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional \bmb{???} covariance structure \citep{hefley2017basis}. 
Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations \bmb{???}.
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we included the additional convergence criterion of effective sample size (ESS) $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}. 
For each replication, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We then compute frequentist coverage to assess the quality of the confidence intervals.
Coverage refers to the proportion of simulations in which the computed confidence intervals include the true values of parameters.
We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
% \bmb{clarify: did you use profile if available, otherwise Wald? You don't give any more detail in your coverage plots. [How different were Wald and profile for these results, or did you not compare them?]}
We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.

@@

Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide when it is best to use phylogenetic random effects, simple grouping, or both. 
Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ will be sufficient. 
However, if multiple observations are available per species, then these simple methods may confound tip variation with residual variation.
In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.
This is equivalent to assuming homoscedasticity using inverse-variance weights for unbalanced datasets. 
Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.

It may be easier to be conservative and include both phylogenetic random effects and grouping random effects
\bmb{clarify? ``iid random effects''?}
but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}). In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.

>>>>

The duality of $\bSigma$ and $\bZ$ for equivalent model specifications is useful for efficient implementations of models that contain a functional (e.g., autoregressive) covariance structure \citep{hefley2017basis}. 
Similarly, we can construct phylogenetic correlations (PC) through $\bZ$.
Suppose that the evolution follows a Brownian-motion process, i.e., continuous traits evolve independently, following an unbiased, continuous-time random walk, along each branch of the phylogeny.
In this case, the phylogenetic variability of a particular species can be written as the sum of the variances of evolutionary changes that occurred on all of the branches in its history. 
Thus, modeling the evolutionary history of each species with a sequence of independent errors with species--branch matrix $\bS$ is equivalent to imposing a correlation $\bC$.
For example, for the phylogeny in figure \ref{fig:tree}, the corresponding $\bS$ takes the form:

@@

Using the R package \pkg{ape} \citep{ape}, we simulated 100 random phylogenetic trees  for each sample size (n = 25, 50, 100 and an additional n = 500 for the multi-group model) and then simulated the responses for each tree using (\ref{eq:single_glmm}, \ref{eq:multiple_glmm}). 
Each realization was fitted using all model variants. 
All simulation parameters are shown in Figure \ref{ssplot} and Figure \ref{msplot}. 
Correlation $\rho_{xy}$ is used instead of covariances $\Sigma_{xy}$ (i.e., $\rho_{xy} = \frac{\Sigma_{xy}}{\Sigma_{x}\Sigma_{y}}$) in the simulations.
%\bmb{???}
%mli{Just to match what lme4/glmmtmb outputs.}
Table \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian GLMM, we included only realizations where we require values of the Gelman-Rubin statistic less than the recommended threshold 1.1. Based on recent concerns about Gelman-Rubin thresholds \citep{vats2018revisiting}), we included the additional convergence criterion of effective sample size (ESS) $>1000$ for the fixed effect parameters ($\beta_{0}$ and $\beta_{1}$) for each replication \citep{vehtari2019rank}. 
For each replication, we sample two chains starting with 10000 iterations. 
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms) to quantify bias and variance (i.e., quality of the point estimate).
We then compute frequentist coverage to assess the quality of the confidence intervals.
Coverage refers to the proportion of simulations in which the computed confidence intervals include the true values of parameters.
We used 95\% Wald confidence intervals for deterministic methods and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
% \bmb{clarify: did you use profile if available, otherwise Wald? You don't give any more detail in your coverage plots. [How different were Wald and profile for these results, or did you not compare them?]}
We also compare computational speed between different platforms to evaluate the efficiency of different platforms and methods.

@@

Nevertheless, for PGLMMs, it is hard to account for all forms of complexities and decide when it is best to use phylogenetic random effects, simple grouping, or both. 
Optimal model complexity depends on experimental design and whether the data provides enough signal to estimate these different levels of variation, which can be strongly confounded.
For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ will be sufficient. 
However, if multiple observations are available per species, then these simple methods may confound tip variation with residual variation.
In this case, multiple observations can summarized to a single measurement (for example, weighted-mean) per species to avoid confounding residual and tip variation.
This is equivalent to assuming homoscedasticity using inverse-variance weights for unbalanced datasets. 
Alternatively, when the within-species variance is actually of interest, accounting for within-species variation (i.e., adding species-level random effects in our example) can automatically handle multiple observations per species.
It may be easier to be conservative and include both phylogenetic random effects and iid grouping random effects by simplifying the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}). 
In this case, the PGLMM collapses to a standard random-slopes model where the group effects are considered to be independent and identically distributed.
